a---
title: "BIO331 -- Lab 07: Bioinformatics and Working with Strings of Biological Data"
subtitle: "Bioinformatics"
author: Dee Ruttenberg
date: Mar. 18, 2026
output: html_document
---

Today we're going to apply our work on R to Bioinformatic data sets. The typical output of a DNA sequencing experiment is a fastq file, consisting of often millions of configs which need to be stitched together (do you remember the two ways we do that?). Fastq files are different from fasta files as they contain quality scores -- this is useful to see how reliable your fastq data is.  The first tool we are going to use is Fastqc, to interpret the results of our fastq files.

# Quality Control

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Rqc")
```

We will use a specific small dataset used specifically for teaching bioinformatics. Rqc will output an .html file (essentially a web page).  For this class, I will ask that you understand specifically the results of 'Average Quality', 'Cycle-Specific Average Quality', 'Cycle-specific Quality Distribution', and 'Cycle-Specific Base Call Proportion'.

```{r}
library(Rqc)
folder <- system.file(package="ShortRead", "extdata/E-MTAB-1147")
rqc(path = folder, pattern = ".fastq.gz")
```

# Bioconductor

Bioconductor and Biostrings are especially powerful tools to examine fasta files in R.  In this section, we'll learn some tools to manipulate them.  

```{r}
library(Biostrings)
seqs <- readDNAStringSet("Full_cmv_259sequences_Aln.fasta")
```

As Fasta files are labeled, you can identify specific sequences, either by loading a specfic label or identifying a keyword:

```{r}
my_seq <- seqs["Europe_United Kingdom_GQ221974.1_Human_herpesvirus_5_strain_3157_complete_genome"]
names_with_uk <- grep("United Kingdom", names(seqs), ignore.case = TRUE)
UK_seqs <- seqs[names_with_uk]
UK_subset <- subseq(UK_seqs, start = 20000, end = 25000)
```

You can, using R, code an aligned genome as a matrix, in order to identify snps. 

```{r}
mat <- as.matrix(UK_seqs)
snp_positions <- which(apply(mat, 2, function(col) {
  bases <- col[col %in% c("A","C","G","T")]
  length(unique(bases)) > 1
}))
snp_matrix <- mat[, snp_positions]
snp_list <- lapply(snp_positions, function(pos) {
  data.frame(
    Sequence = rownames(mat),
    Position = pos,
    Base = mat[, pos]
  )
})
names(snp_list) <- paste0("Pos_", snp_positions)
```

#Alignment

The final skill we need to learn is how to align .fasta files to a reference genome.  The data we have today is aligned so we'll have to arbitrarily "unalign" it.  We will use our first sequence as the reference.  

```{r}
UK_seqs_nogap <- DNAStringSet(gsub("-", "", as.character(UK_seqs))) 
ref <- seqs[6]
```

The region of interest from hcmv will be our query site.   

```{r}
library(Biostrings)
query <- readDNAStringSet("example.fasta")
ref <- seqs["Europe_Germany_JX512204.1_Human_herpesvirus_5_strain_HAN16_complete_genome"]
ref <- readDNAStringSet("hcmvgenome.fasta")
query <- DNAString(toupper(as.character(query)))
ref <- DNAStringSet(gsub("-", "", as.character(ref))) 

hits <- matchPattern(query, ref[[1]], max.mismatch=20)

```

We can now look at our aligned data!

```{r}
hits
```

# Homework Question 1: Interpretting QC data.

a. Look at the 'Cycle-specific Base Call Proportion' rqc file.  What is the ratio of A/C/G/T in cycles 1-10. What is the ratio of A/C/G/T in cycles 65-75?  Based on your knowledge of DNA-sequencing, propose a hypothesis for why the cycles have a different ratio.

## The ratio of A/C/G/T seems to flucuate in cycles 1-10, whereas in cycles 50 on, the ratio stays pretty steady (about 25% of each nucleotide). My guess as to why the early cycles flucuate while the later cycles have a steady ratio is that maybe the primers used to begin sequencing weren't removed, so the nucleotides from the primers are being counted in the beginning cycles which would result in weird ratios between the nucleotides. 

b. Look at the 'Cycle specific Quality Distribution'. What is the typical quality of cycles 1-10?  What is the typical quality of cycles 65-75?  Based on your knowledge of DNA-sequencing, propose a hypothesis for why the cycles have a different quality

## The quality distribution graphs show that the first ten cycles are of higher quality, while the last ten cycles are of lower quality. This is likely because you are using a reversible primer, so if it makes an error in an earlier cycle, you will just continue to build on that error in all of the following cycles, meaning the quality and correctness of the sequencing will very much decrease. 

# Homework Question 2: Loading a File, Reverse Complements

For one of the research projects, we will be studying a specific protein coding site of hcmv associated with the transition of hcmv from latent to active. Align the following two queries to any genome in the reference file. Which one successfully aligns? What does that mean about the site of interest?

```{r}
query <- readDNAStringSet("hcmv15-hcmv1b-1.fasta")
reversecompquery <- reverseComplement(query)
ref <- seqs["Europe_Germany_JX512204.1_Human_herpesvirus_5_strain_HAN16_complete_genome"]
query <- DNAString(toupper(as.character(query)))
ref <- DNAStringSet(gsub("-", "", as.character(ref)))

qhits <- matchPattern(query, ref[[1]], max.mismatch=20)

reversecompquery <- DNAString(toupper(as.character(reversecompquery)))

reverseqhits <- matchPattern(reversecompquery, ref[[1]], max.mismatch = 20)


qhits
reverseqhits


## The reverse query successfully aligns, which means that the site of interest is reading in the reverse (from 5' -> 3' rather than 3' -> 5')
```

# Homework Question 3: Isolating a Site

Isolate the 'hcmv15-hcmv1b-1.fasta' site in all 259 seqs. Your final result should be a DNAString set of 259 sequences about 4-5 kb long.  

```{r}
UK_subset <- subseq(UK_seqs, start = 187075, end = 191699)

print(UK_subset)
print(seqs)

library(pwalign)

start_pos = start(reverseqhits)
end_pos   = end(reverseqhits)

seqs_nogap = DNAStringSet(gsub("-", "", as.character(seqs)))

site_hits <- lapply(seqs_nogap, function(s) {
  hits <- matchPattern(reversecompquery, s, max.mismatch = 20)
  if (length(hits) > 0) {
    as.character(subseq(s, start = start(hits)[1], end = end(hits)[1]))
  } else {
    NULL
  }
})

site_hits_clean <- Filter(Negate(is.null), site_hits)

char_vec <- unlist(site_hits_clean)
char_vec <- char_vec[!is.na(char_vec) & nchar(char_vec) > 0]

result_259 <- DNAStringSet(char_vec)

result_259
length(result_259)
width(result_259)

```